### 2024-02-14 parts/intro.tex
uploading to ICALP

Traditional results from automata theory state the equivalence of languages
defined in terms of deterministic finite automata and regular expressions
\cite{kleene1956representation}, Monadic Second Order Logic ($\MSO$)
\cite{buchi1960weak}, or finite monoids \cite{schutzenberger1961definition}.
Among regular languages, there exists a robust subclass sharing similar
characterizations, namely the \emph{star-free} regular languages. These are
obtained via counter-free (minimal) automata \cite{mcnaughton1971counter},
represented by star-free regular expressions \cite{schutzenberger1965finite},
definable in First Order Logic ($\FO$) \cite{perrin1986first}, and recognized
by aperiodic monoids \cite{schutzenberger1965finite}. It is decidable whether a
regular language is aperiodic, thanks to the characterizations in terms of
minimal automaton and/or syntactic monoid.

There have been recurrent attempts to generalize the notions of regularity from
languages to functions, leading to classes of various expressiveness
(sequential, rational, regular, and \kl{polyregular functions}). In this paper,
we will focus on \kl{polyregular functions} ($\Poly$), that shares with regular
languages its characterizations in terms of logic (via $\MSO$ interpretations),
automata (via pebble automata), and adds new bridges with programming languages
via its relationships with lambda-terms and so-called for-transducers
\cite{bojanczyk2019string}. As for regular languages, the class
$\SF$ of \intro{star-free polyregular function} can equivalently be introduced
in terms of first-order definability, counter-free pebble transducers, and
suitable restrictions of for-tranducers.

Classical problems, including the decidability of aperiodicity, become
difficult in the case of \kl{polyregular functions}, due to the lack of
canonical objects akin to the syntactic monoid or the minimal automaton.
Indeed, for sequential functions that admit a canonical object, aperiodicity is
decidable \cite{choffrut03}, and more recently, aperiodicity has been proven
decidable for rational functions
\cite{filiot2016aperiodicity,filiot2018canonical}. The problem remains open for
regular functions, although it has been proven decidable when changing the
semantics of function equality to account for the \emph{origins} in the output
\cite{bojanczyk14}. To construct subclasses of \kl{polyregular function}, an
interesting parameter is the \intro{growth rate} of a function $f$, which is given
by a number $k \in \Nat$, which can be used to create a strict hierarchy inside
$\Poly$. We write $\Poly[k]$ for the \kl{polyregular functions} of \kl{growth
rate} $k$. It has been shown that the parameter $k$ controls the maximal
complexity of the $\MSO$ interpretation needed to represent the function
\cite{bojanczyk2022transducers}.

Another way to approach the problem of deciding if a \kl{polyregular function}
is \kl{star-free}, is to not restrict the computational power (via the
\kl{growth rate}), nor change the semantics (by adding origins), but consider
restrictions of the output alphabet. In particular, there have been successful
results proven on \emph{unary polyregular functions}, also known as
\kl{$\Nat$-polyregular functions} ($\NPoly$)
\cite{doueneau2021pebble,doueneau2022hiding}. For instance, the \kl{growth
rate} $k$ of a function $f \in \NPoly$ provides the minimal number of pebbles
needed to compute $f$ \cite{doueneau2021pebble}, which does not hold on $\Poly$
\cite{bojanczyk2022transducers}. Let us remark that the restriction to
\emph{unary output} in the case of \kl{polyregular functions} can be lifted to
outputs in the free commuative semigroup generated by a finite alphabet
$\Gamma$. In this sense, the shift to \emph{unary output} is really about
enforcing commutativity on the output.

Recently, aperiodicity has been shown decidable over a restricted class of
\kl{polyregular functions} where the ouptut belongs to $\Rel$ instead of
$\Nat$, that are called \kl{$\Rel$-polyregular functions} ($\ZPoly$)
\cite{LOPEZ23b}. These functions generalize \kl{$\Nat$-polyregular functions}
by allowing negative outputs, and therefore enabling a limited form of
backtracking. As for the unary outputs, these results generalize to
\kl{polyregular functions} having outputs in the free commutative \emph{group}
generated by a finite alphabet $\Gamma$. The decidability of aperiodicity for
\kl{$\Rel$-polyregular functions} relies on the construction of a canonical
object called the \emph{residual transducer}, the latter being essentially
based on differences between functions, crucially leveraging negative outputs
(i.e., the presence of inverse elements). If the proof method does not carry
from $\Rel$-output functions to $\Nat$-output functions, it was conjectured
that the semantic characterization of \kl{star-free $\Rel$-polyregular
functions} ($\ZSF$) would continue to characterize \kl{star-free
$\Nat$-polyregular functions} ($\NSF$).

@@

It is well known that $\NPoly$ (resp. $\ZPoly$) is proper subclass respectively
of \kl{$\Nat$-rational series} ($\NRat$) (resp. of \kl{$\Rel$-rational series}
$\ZRat$), that have been characterized in terms of \kl{growth rate}, ambiguity,
and a notion of \emph{quantitative} $\MSO$
\cite{schutzenberger1962,kreutzer2013}, and therefore can benefit from generic
theorems obtained on these \cite{berstel2011noncommutative}. However, for these
larger classes of functions, several competing notions of aperiodicity exist
\cite{droste2019aperiodic,reutenauer_series_1980,LOPEZ23b}, and the
decidability of the ones that extends $\ZSF$ and $\NSF$ remains open.
Furthermore, the decidability of $\NRat$ inside $\ZRat$ is also a longstanding
open problem, that was to the knowledge of the authors, only solved by
\textcite{KARH77} in the case of \kl{$\Rel$-rational series} that compute
\emph{polynomials}, with an incorrect proof. The decidability of $\NPoly$
inside $\ZPoly$ (resp. $\NSF$ inside $\ZSF$) remains an open problem \cite[Open
question 5.55]{gaetanphd}.

@@

\begin{figure}
    \includestandalone[width=5cm]{tikz/class-inclusions}
    \caption{
        Decidability and inclusions of classes of functions,
        arranged along two axes. The first one is the complexity
        of the output alphabet ($\Nat$, $\Rel$, $\Sigma$). The second
        one is the allowed computational power
        (star-free polyregular functions, polyregular functions, 
        rational series).
    }
    \label{previously-known-inclusions:fig}
\end{figure}

@@

Finally, we make a first step towards understanding the non-commutative by
introducing a canonical model of computation for \kl{$\Nat$-polyregular
functions} built upon residuals, called the \kl{residual transducer}. This
allows the characterization of $\NPoly[k]$ as the class of functions $f$ such
that the collection $\Res(f)$ of all \kl{residuals} of $f$ is a
\kl{well-quasi-ordering}, generalizing the characterization of $\ZPoly[k]$ as
the class of functions $f$ such that $\Res(f)$ has finite index (for some
equivalence relation). The \kl{residual transducer} is effectively computable
if $f$ is a \kl{commutative} \kl{$\Nat$-polyregular function}, and it is
conjectured to be computable for all functions in $\NPoly$.

\paragraph*{Outline of the paper.} In \cref{preliminaries:sec}, we introduce
the multiple characterizations of \kl{$\Nat$-polyregular function} (resp.
$\Rel$-polyregular functions), show that one can decide if a function $f \in
\ZPoly$ is \kl{commutative} (\cref{decidable-commutative-poly:lemma}), and give the
basic definitions used on multivariate polynomials. In \cref{polynomials:sec},
we focus on \emph{polynomials}, by introducing the \cref{karh:thm} of
\textcite{KARH77}, providing a counter-example in \cref{thm:counter-example},
and an alternative characterization in \cref{corrected-version:thm}. In
\cref{beyond-polynomials:sec}, we leverage the results over polynomials to
decide $\NPoly$ inside $\ZPoly$ in the \kl{commutative} case
(\cref{decidable-n-poly:thm}), and prove that \cref{zsf-nsf:conjecture} holds
for \kl{commutative} functions (\cref{zsf-npoly-nsf:thm}). Finally, in
\cref{beyond-commutative:sec}, we introduce the notion of \kl{residual
transducer} (\cref{residual-transducer:def}), and prove in
\cref{non-commutative-npoly:thm} that it characterizes $\NPoly$ inside
$\ZPoly$.

>>>>

Traditional results from automata theory state the equivalence of languages
defined in terms of deterministic finite automata and regular expressions
\cite{kleene1956representation}, Monadic Second Order Logic ($\intro*\MSO$)
\cite{buchi1960weak}, or finite monoids \cite{schutzenberger1961definition}.
Among regular languages, there exists a robust subclass sharing similar
characterizations, namely the \emph{star-free} regular languages. These are
obtained via counter-free (minimal) automata \cite{mcnaughton1971counter},
represented by star-free regular expressions \cite{schutzenberger1965finite},
definable in First Order Logic ($\intro*\FO$) \cite{perrin1986first}, and recognized
by aperiodic monoids \cite{schutzenberger1965finite}. It is decidable whether a
regular language is aperiodic, thanks to the characterizations in terms of
minimal automaton and/or syntactic monoid.

There have been recurrent attempts to generalize the notions of regularity from
languages to functions, leading to classes of various expressiveness
(sequential, rational, regular, and \kl{polyregular functions}). In this paper,
we will focus on \intro{polyregular functions} ($\intro*\Poly$), that shares with regular
languages its characterizations in terms of logic (via $\MSO$ interpretations),
automata (via pebble automata), and adds new bridges with programming languages
via its relationships with lambda-terms and so-called for-transducers
\cite{bojanczyk2019string}. As for regular languages, the class
$\intro*\SF$ of \intro{star-free polyregular function} can equivalently be introduced
in terms of first-order definability, counter-free pebble transducers, and
suitable restrictions of for-tranducers.

Classical problems, including the decidability of aperiodicity, become
difficult in the case of \kl{polyregular functions}, due to the lack of
canonical objects akin to the syntactic monoid or the minimal automaton. For
sequential functions (that admit a canonical object), aperiodicity is decidable
\cite{choffrut03}, and more recently, aperiodicity has been proven decidable
for rational functions \cite{filiot2016aperiodicity,filiot2018canonical}. The
problem remains open for regular functions, although it has been proven
decidable when changing the semantics of function equality to account for the
\emph{origins} in the output \cite{bojanczyk14}. To construct subclasses of
\kl{polyregular function}, an interesting parameter is the \intro{growth rate}
of a function $f$, which is given by a number $k \in \Nat$, which can be used
to create a strict hierarchy inside $\Poly$. All functions in $\Poly$ have
\intro{polynomial growth}, and we can write $\Poly[k]$ for the \kl{polyregular
functions} of \kl{growth rate} $k$. It has been shown that the parameter $k$
controls the maximal complexity of the $\MSO$ interpretation needed to
represent the function \cite{bojanczyk2022transducers}.

Another way to approach the problem of deciding if a \kl{polyregular function}
is \kl{star-free}, is to not restrict the computational power (via the
\kl{growth rate}), nor change the semantics (by adding origins), but consider
restrictions of the output alphabet. In particular, there have been successful
results proven on \emph{unary polyregular functions}, also known as
\kl{$\Nat$-polyregular functions} ($\intro*\NPoly$)
\cite{doueneau2021pebble,doueneau2022hiding}. For instance, the \kl{growth
rate} $k$ of a function $f \in \NPoly$ provides the minimal number of pebbles
needed to compute $f$ \cite{doueneau2021pebble}, which does not hold on $\Poly$
\cite{bojanczyk2022transducers}. Let us remark that the restriction to
\emph{unary output} in the case of \kl{polyregular functions} can be lifted to
outputs in the free commuative semigroup generated by a finite alphabet
$\Gamma$. In this sense, the shift to \emph{unary output} is really about
enforcing commutativity on the output.

Recently, aperiodicity has been shown decidable over a restricted class of
\kl{polyregular functions} where the ouptut belongs to $\Rel$ instead of
$\Nat$, that are called \kl{$\Rel$-polyregular functions} ($\intro*\ZPoly$)
\cite{LOPEZ23b}. These functions generalize \kl{$\Nat$-polyregular functions}
by allowing negative outputs, and therefore enabling a limited form of
backtracking. As for the unary outputs, these results generalize to
\kl{polyregular functions} having outputs in the free commutative \emph{group}
generated by a finite alphabet $\Gamma$. The decidability of aperiodicity for
\kl{$\Rel$-polyregular functions} relies on the construction of a canonical
object called the \emph{residual transducer}, the latter being essentially
based on differences between functions, crucially leveraging negative outputs
(i.e., the presence of inverse elements). Although the proof method does not carry
from $\Rel$-output functions to $\Nat$-output functions, it was conjectured
that the semantic characterization of \kl{star-free $\Rel$-polyregular
functions} ($\intro*\ZSF$) would continue to characterize \kl{star-free
$\Nat$-polyregular functions} ($\intro*\NSF$).

@@

It is well known that $\NPoly$ (resp. $\ZPoly$) is proper subclass respectively
of \intro{$\Nat$-rational series} ($\intro*\NRat$) (resp. of \intro{$\Rel$-rational
series} $\intro*\ZRat$), that have been characterized in terms of \kl{growth rate},
ambiguity, and a notion of \emph{quantitative} $\MSO$
\cite{schutzenberger1962,kreutzer2013}, and therefore can benefit from generic
theorems obtained on these \cite{berstel2011noncommutative}. However, for these
larger classes of functions, several competing notions of aperiodicity exist
\cite{droste2019aperiodic,reutenauer_series_1980,LOPEZ23b}, and the
decidability of the ones that extends $\ZSF$ and $\NSF$ remains open.
Furthermore, the decidability of $\NRat$ inside $\ZRat$ is also a longstanding
open problem, that was to the knowledge of the authors, only solved by
\textcite{KARH77} in the case of \kl{$\Rel$-rational series} that compute
\emph{polynomials}, with an incorrect proof. Furthermore, decidability of
$\NPoly$ inside $\ZPoly$ (resp. $\NSF$ inside $\ZSF$) remains an open problem
\cite[Open question 5.55]{gaetanphd}.

@@

\begin{figure}
    \centering
    \includestandalone[width=7cm]{tikz/class-inclusions}
    \caption{
        Decidability and inclusions of classes of functions,
        arranged along two axes. The first one is the complexity
        of the output alphabet ($\Nat$, $\Rel$, $\Sigma$). The second
        one is the allowed computational power
        (star-free polyregular functions, polyregular functions, 
        rational series).
    }
    \label{previously-known-inclusions:fig}
\end{figure}

@@

Finally, we make a first step towards understanding the non-commutative case by
introducing a canonical model of computation for \kl{$\Nat$-polyregular
functions} built upon residuals, called the \kl{residual transducer}. This
residual transducer is obtained by shifting attention from the notion of
equivalence relations of finite index over $\Sigma^*$, used in traditional
automata theory and in the study of $\ZPoly$, to the notion of
\kl{well-quasi-orderings}. The residual transducer is effectively computable if
$f$ is \kl{commutative} and \kl{$\Nat$-polyregular}, and it is conjectured to
be computable for all functions in $\NPoly$.

\paragraph*{Outline of the paper.} In \cref{preliminaries:sec}, we introduce
the multiple characterizations of \kl{$\Nat$-polyregular function} (resp.
$\Rel$-polyregular functions), show that one can decide if a function $f \in
\ZPoly$ is \kl{commutative} (\cref{decidable-commutative-poly:lemma}), and give
the basic definitions used on multivariate polynomials. In
\cref{polynomials:sec}, we focus on \emph{polynomials}, by introducing the
\cref{karh:thm} of \textcite{KARH77}, providing a counter-example in
\cref{thm:counter-example}, and an alternative characterization in
\cref{corrected-version:thm}. In \cref{beyond-polynomials:sec}, we leverage the
results over polynomials to decide $\NPoly$ inside $\ZPoly$ in the
\kl{commutative} case (\cref{decidable-n-poly:thm}), and prove that
\cref{zsf-nsf:conjecture} holds for \kl{commutative} functions
(\cref{zsf-npoly-nsf:thm}). Finally, in \cref{beyond-commutative:sec}, we
introduce the notion of \kl{residual transducer}
(\cref{residual-transducer:def}), and prove in \cref{non-commutative-npoly:thm}
that it characterizes $\NPoly$ inside $\ZPoly$ through a notion of
\kl{well-quasi-ordered} collection of residuals.

### 2024-02-14 parts/prelim.tex
uploading to ICALP

\AP Let us briefly recall in the following lemma alternative characterizations
of $\NPoly$ (resp. $\ZPoly$), some of which will be helpful in the upcoming
analysis of their \kl{commutative} counterpart, and others that provide
a nicer syntax to construct examples. These characterizations are known, and
can be found in \cite{gaetanphd,LOPEZ23b}.

@@

Similar characterizations can be obtained for $\NSF$ and $\ZSF$.

@@

\AP Inside $\ZPoly$ there is an infinite hierarchy of classes, defined by
limiting the maximal number of free variables used to construct the function as
a sum of counting formulas (\cref{npoly-counting-mso:item}). We write
$\ZPoly[k]$ (resp. $\NPoly[k]$, $\ZSF[k]$, $\NSF[k]$) for the functions in
$\ZPoly$ (resp. $\NPoly$, $\ZSF$, $\NSF$) that can be constructed using at most
$k$ free variables. It has been proven in
\cite{LOPEZ23b,bojanczyk2022transducers} that the classes $\ZPoly[k]$ (resp.
$\NPoly[k]$, $\ZSF[k]$, $\NSF[k]$) are characterized by the \kl{growth rate},
that is, $f \in \ZPoly[k]$ if and only if $f \in \ZPoly$ and $\growth{f}(n) =
\bigO(n^k)$, and similarly for the other classes.

@@

\AP Because this paper focuses on functions that are commutative both in the
input and the output, let us introduce precisely what it means and how it
relates to other functions from $\Nat^k \to \Nat$. For that, let $\Sigma$ be a
finite alphabet. We define $\commute \colon \Sigma^* \to \Nat^\Sigma$ for the
map that counts occurrences of every letter in the input word, namely: $
\commute[w] \defined (a \mapsto \card[a]{w})$. Let $X$ be a set. A function $f
\colon \Sigma^* \to X$ is \intro{commutative} whenever for all $u,v \in
\Sigma^*$, $f(uv) = f(vu)$. Equivalently, it is \reintro{commutative} whenever
there exists a map $g \colon \Nat^\Sigma \to X$ such that $g \circ \commute =
f$.

@@

Let us briefly remark that the characterizations of subclasses of
\kl{polyregular functions} that are obtained in practice can be obtained by
pre-composition with a \kl{commutative} \kl{polyregular function}. While the
two following
\cref{pre-compose-growth-commut:lemma,pre-compose-sf-commut:lemma} do not
provide simpler proofs, they suggest the central position of \kl{commutative}
\kl{star-free polyregular functions} as \emph{test functions} for various
properties (growth rate, aperiodicity).

\begin{lemma}
    \label{pre-compose-growth-commut:lemma}
    Let $f \in \ZRat$, and $k \in \Nat$,
    the following are equivalent:
    \begin{enumerate}
        \item $f \in \ZPoly[k]$,
        \item For every \kl{commutative} \kl{star-free polyregular function} $h$
            of \kl{growth rate} $l \in \Nat$,
            $(f \circ h) \in \ZPoly[k+l-1]$.
    \end{enumerate}
\end{lemma}

\begin{lemma}
    \label{pre-compose-sf-commut:lemma}
    Let $f \in \ZPoly$.
    The following are equivalent:
    \begin{enumerate}
        \item $f \in \ZSF$,
        \item For every \kl{commutative} \kl{star-free polyregular function} $h$,
            $(f \circ h) \in \ZSF$.
    \end{enumerate}
\end{lemma}

@@

\AP Let us now give some vocabulary on polynomials with multiple indeterminate
over $\Rel$. A \intro{monomial} is a product of indeterminates and integers.
For instance, $XY$ is a \kl{monomial}, $3 X$ is a \kl{monomial}, $-Y$ is a
\kl{monomial}, but $X + Y$ or $2X^2 + XY$ are not. We write $\Monomials[X_1,
\dots, X_n]$ for the set of \kl{monomials} over these indeterminates.
Every polynomial $P \in \Rel[X_1, \dots, X_n]$ decomposes uniquely
into a sum of \kl{monomials}.

\AP A \kl{monomial} $S$ \intro{divides} a \kl{monomial} $T$, when $S$ divides
$T$ seen as polynomials in $\Rat$. For instance, $2X$ \kl{divides} $XY$, $-YZ$
\kl{divides} $X^2 Y Z^3$, and $Y$ does not \kl{divide} $X$. In the
decomposition of $P \in \Rel[X_1, \dots, X_k]$, a \kl{monomial} is a
\intro{maximal monomial} if it is a maximal element for the \kl{divisibility
ordering} of \kl{monomials}. In the polynomial $P(X,Y) \defined X^2 - 2XY + X^2
+ X + Y$, the set of \reintro{maximal monomials}, written
$\MaximalMonomials(P)$, is $\set{X^2,  -2 XY,  X^2}$. A \kl{monomial} has a
\intro{non-negative coefficient} if its multiplicative constant belongs to
$\Nat$, or equivalently, if it is \kl{non-negative} as a polynomial. In the
polynomial $P(X,Y) \defined (X - Y)^2$, the \kl{non-negative} \kl{monomials}
are $X^2$ and $Y^2$.

>>>>

\AP Let us briefly recall in the following lemma alternative characterizations
of $\NPoly$ (resp. $\ZPoly$), some of which will be helpful in the upcoming
analysis of their \kl{commutative} counterpart, and others that provide
a nicer syntax to construct examples. These characterizations were previously known, and
can be found in \cite{gaetanphd,LOPEZ23b}.

@@

Similar characterizations can be obtained for $\NSF$ and $\ZSF$, which correctly
generalizes aperiodicity to functions as \cref{regular-language:ex} illustrates.

@@

\AP Inside $\ZPoly$ there is an infinite hierarchy of classes, defined by
limiting the maximal number of free variables used to construct the function as
a sum of counting formulas (\cref{npoly-counting-mso:item}). We write
$\ZPoly[k]$ (resp. $\NPoly[k]$, $\ZSF[k]$, $\NSF[k]$) for the functions in
$\ZPoly$ (resp. $\NPoly$, $\ZSF$, $\NSF$) that can be constructed using at most
$k$ free variables. It has been proven in
\cite{LOPEZ23b,bojanczyk2022transducers} that the classes $\ZPoly[k]$ (resp.
$\NPoly[k]$, $\ZSF[k]$, $\NSF[k]$) are characterized by the \kl{growth rate},
that is, $f \in \ZPoly[k]$ if and only if $f \in \ZPoly$ and $\card{f(w)} =
\bigO\left(\card{w}^k\right)$. and similarly for the other classes.

@@

\AP Because this paper focuses on functions that are commutative both in the
input and the output, let us introduce precisely what it means and how it
relates to other functions from $\Nat^k \to \Nat$. For that, let $\Sigma$ be a
finite alphabet. We define $\commute \colon \Sigma^* \to \Nat^\Sigma$ for the
map that counts occurrences of every letter in the input word (that is,
computes the Parikh vector) namely: $ \commute[w] \defined (a \mapsto
\card[a]{w})$. Let $X$ be a set, a function $f \colon \Sigma^* \to X$ is
\intro{commutative} whenever for all $u,v \in \Sigma^*$, $f(uv) = f(vu)$.
Equivalently, it is \reintro{commutative} whenever there exists a map $g \colon
\Nat^\Sigma \to X$ such that $g \circ \commute = f$.

@@

@@

\AP Let us now give some vocabulary on polynomials with multiple indeterminate
over $\Rel$. A \intro{monomial} is a product of indeterminates and integers.
For instance, $XY$ is a \kl{monomial}, $3 X$ is a \kl{monomial}, $-Y$ is a
\kl{monomial}, but $X + Y$ or $2X^2 + XY$ are not. We write $\intro*\Monomials[X_1,
\dots, X_n]$ for the set of \kl{monomials} over these indeterminates.
Every polynomial $P \in \Rel[X_1, \dots, X_n]$ decomposes uniquely
into a sum of \kl{monomials}.

\AP A \kl{monomial} $S$ \intro{divides} a \kl{monomial} $T$, when $S$ divides
$T$ seen as polynomials in $\Rat$. For instance, $2X$ \kl{divides} $XY$, $-YZ$
\kl{divides} $X^2 Y Z^3$, and $Y$ does not \kl{divide} $X$. In the
decomposition of $P \in \Rel[X_1, \dots, X_k]$, a \kl{monomial} is a
\intro{maximal monomial} if it is a maximal element for the \kl{divisibility
ordering} of \kl{monomials}. In the polynomial $P(X,Y) \defined X^2 - 2XY + X^2
+ X + Y$, the set of \reintro{maximal monomials}, written
$\intro*\MaximalMonomials(P)$, is $\set{X^2,  -2 XY,  X^2}$.  In the polynomial
$P(X,Y) \defined (X - Y)^2$, the \kl{non-negative} \kl{monomials} are $X^2$ and
$Y^2$.

### 2024-02-14 parts/starfree.tex
uploading to ICALP

Let us now go beyond polynomials and consider the case of \kl{commutative}
\kl{$\Rel$-polyregular functions}. The hypothesis of \kl{commutativity} reduces
$\ZPoly$ to a combination of a finite number of polynomials, arranged by
counting modulo some numbers. This characterization will be proven in
\cref{decompose-polynomial:lem}, and will be the key ingredient in the decision
procedures of $\NPoly$ inside $\ZPoly$, and $\NSF$ inside $\NPoly$. Note that
\cref{decompose-polynomial:lem} is actually a refinement of
\cref{n-poly-combinatorics:lem} under the extra assumption of
\kl{commutativity}.

\AP To formally state the lemma, we will need some notations that will
formalize the intuitive idea of counting tuples of integers modulo some
constant. Let $\omega \in \Nat$, we write the set of \intro{types modulo
$\omega$} $\ModuloTypes[\omega] \defined \set{0, \dots, \omega^2} \times \set{0,
    \dots, \omega - 1}$. To a number $n \in \Nat$, we associate its
    \intro{$\omega$-type} written $\moduloType[\omega](n)$ which is defined as the pair
    $(t, r)$ where $t = \min (n, \omega^2)$ and $r$ is the rest of the Euclidian
    division of $n$ by $\omega$. This notion of type is lifted to vectors in
    $\Nat^p$ for any $p \in \Nat$ by pointwise application.

@@

The goal of this section is to go beyond polynomials and characterize all
\kl{$\Rel$-polyregular functions} that are \kl{$\Nat$-polyregular}, under the
assumption that they are \kl{commutative}. We will provide a semantic property
in the form of \cref{k-combinatorial:def}, that is conjecture to characterize
$\NPoly$ inside $\ZPoly$, even in the non-commutative setting.

@@

Let us now prove that the above characterizations of \kl{commutative}
\kl{$\Nat$-polyregular functions} can be combined with the recent advances in
the study of \kl{$\Rel$-polyregular functions} \cite{LOPEZ23b} to prove the
following theorem. As a corollary of \cref{zsf-npoly-nsf:thm} and the effective
decision procedures of $\ZSF$ obtained in \cite{LOPEZ23b}, we conclude that
$\NSF$ is decidable inside $\NPoly$ under the assumption of commutativity, and
that the effective procedure can be used The key ingredient of this section is
the use of a semantic characterization of \kl{star-free $\Rel$-polyregular
functions} among \kl{$\Rel$-rational series} that generalizes the aperiodicity
of languages to functions by the means of polynomial behaviors.

@@

\begin{theorem}
    \label{zsf-npoly-nsf:thm}
    Let $\Sigma$ be a finite alphabet, $k \in \Nat$,
    and $f \colon \Sigma^* \to \Rel$ be a \kl{commutative}
    \kl{$\Nat$-polyregular function}.
    Then, the following are equivalent:
    \begin{itemize}
        \item $f$ is \kl{ultimately polynomial},
        \item $f \in \ZSF$,
        \item $f \in \NSF[k]$.
    \end{itemize}
    Furthermore, membership is decidable and effective conversions
    are possible.
\end{theorem}

>>>>

The goal of this section is to go beyond polynomials and consider the case of
\kl{commutative} \kl{$\Rel$-polyregular functions}. The hypothesis of
\kl{commutativity} reduces $\ZPoly$ to a combination of a finite number of
polynomials, arranged by counting modulo some numbers. This characterization
will be proven in \cref{decompose-polynomial:lem}, and will be the key
ingredient in the decision procedures of $\NPoly$ inside $\ZPoly$, and $\NSF$
inside $\NPoly$. Note that \cref{decompose-polynomial:lem} is actually a
refinement of \cref{n-poly-combinatorics:lem} under the extra assumption of
\kl{commutativity}.

\AP To actually state the lemma, we will need some notations that will
formalize the intuitive idea of counting tuples of integers modulo some
constant. Let $\omega \in \Nat$, we define the set $\intro*\ModuloTypes[\omega]
\defined \set{0, \dots, \omega^2} \times \set{0, \dots, \omega - 1}$ of
\intro{types modulo $\omega$}. To a number $n \in \Nat$, we associate its
\reintro{$\omega$-type} written $\intro*\moduloType[\omega](n)$ which is
defined as the pair $(t, r)$ where $t = \min (n, \omega^2)$ and $r$ is the rest
of the Euclidian division of $n$ by $\omega$. This notion of type is lifted to
vectors in $\Nat^p$ for any $p \in \Nat$ by pointwise application.

@@

In the spirit of previous characterizations of subclasses of $\ZPoly$ in terms
of \emph{polynomial pumping arguments}
\cite{doueneau2021pebble,doueneau2022hiding,LOPEZ23b}, we will provide a
semantic property in the form of \cref{k-combinatorial:def}, that is conjecture
to characterize $\NPoly$ inside $\ZPoly$, even in the non-commutative setting.

@@

Let us now prove that the above characterizations of \kl{commutative}
\kl{$\Nat$-polyregular functions} can be combined with the recent advances in
the study of \kl{$\Rel$-polyregular functions} \cite{LOPEZ23b} to decide
aperiodicity. The key ingredient of this study is the use of a semantic
characterization of \kl{star-free $\Rel$-polyregular functions} among
\kl{$\Rel$-rational series} that generalizes the aperiodicity of languages to
functions by the means of polynomial behaviors.

@@

\begin{theorem}
    \label{zsf-npoly-nsf:thm}
    Let $\Sigma$ be a finite alphabet, 
    and $f \colon \Sigma^* \to \Rel$ be a \kl{commutative}
    \kl{$\Nat$-polyregular function}.
    Then, the following are equivalent:
    \begin{enumerate}
        \item $f$ is \kl{ultimately polynomial},
        \item $f \in \ZSF$,
        \item $f \in \NSF$.
    \end{enumerate}
    Furthermore, membership is decidable and effective conversions
    are possible.
\end{theorem}
